Change in Mobility and Site Occupation during the Late Pleistocene in Korea

### 2019-11-24 analysis/paper/paper.Rmd
fixed typoes
move Rcode for figures to here. Not everything is successful...

# Introduction

@@

@@

```{r get-data, eval = FALSE}
# Note the path that we need to use to access our data files when rendering this document
my_data <- read.csv(here::here('analysis', 'data', 'raw_data', 'my_csv_file.csv'))
```

@@

@@

@@

@@

Umap

@@

2. Population change
Interpreted in HBE, we expected there might be close connection between environmental change and the appearance of new technology based on the prediction that the new invention was the result of risk. The research suggests that the environmental pressure coule be related to population change. We cannot simply predicate that the increased population around 35ka was due to dispersal of Homo Sapiens since there is no reliable human remains in South Korea. However, the migration of Homo sapiens was not impossible scenario at the same time. In fact, one model explains that the emergence of stemmed points and blade industry came from outside of the Korean peninsula and the new technology was the result of the continuous influx of modern human populations from Siberia while the earlier coarse flake trandition was come from southern China [@Bae_2010]. Since this migration model barely explains about preexisted stone artifact assemblages and people in the Korean Peninsula, it is hard to accept his argument without knowing about the native people. In short, we assume that the stemmed points appeared related to the increased population and it can be related to the modern human dispersal or indigenously increased occupants.

@@

>>>>

```{r setup, include=FALSE}
suppressMessages(library(tidyverse))
library(glue)
library(ggrepel)

# data from PhD data sheet, not KAS sheet.
mydata <- read.csv("~/koreapaleolithicmobilityoccupation/General_info.csv")

# raw material data from KAS data sheet.
kasr <- read.csv("~/koreapaleolithicmobilityoccupation/Rawmaterial_info.csv")

# assemblage composition data from KAS data sheet.
kasa <- read.csv("~/koreapaleolithicmobilityoccupation/Assemblage_info.csv")

#volume of the cultural layer from KAS data sheet.
kasv <- read.csv("~/koreapaleolithicmobilityoccupation/Dating_info.csv")

dates <- read.csv("~/koreapaleolithicmobilityoccupation/Radiocarbon dates.csv")

```
# Introduction

@@

```{r}
#I couldn't find the most recent map that you made. I assume that you don't share the code, but showed me the image so that I just copied it for SAA poster
#This is what I made first before you made the recent one

korean_archaeological_sites <- readxl::read_excel("~/koreapaleolithicmobilityoccupation/korean-archaeologica-sites.xlsx")

library(tidyverse)

coords <- 
	korean_archaeological_sites %>% 
	select(site_name, 
				 east_longitude, 
				 north_latitude) %>% 
	separate(east_longitude, 
					 into = c("e_deg", "e_min", "e_sec"),
					 sep =  "°|\\′") %>% 
	mutate(e_min = str_squish(e_min),
				 e_sec = str_squish(e_sec)) %>% 
	separate(e_min, 
					 into = c("e_min", "e_sec1"), 
					 sep = "'") %>% 
	mutate_at(vars(e_deg, e_min, e_sec, e_sec1), parse_number) %>% 
	mutate(e_sec = ifelse(is.na(e_sec), e_sec1, e_sec)) %>% 
	select(-e_sec1) %>% 
	separate(north_latitude, 
					 into = c("n_deg", "n_min", "n_sec"),
					 sep =  "°|\\′") %>% 
	mutate(n_min = str_squish(n_min),
				 n_sec = str_squish(n_sec)) %>% 
	separate(n_min, 
					 into = c("n_min", "n_sec1"), 
					 sep = "'") %>% 
	mutate_at(vars(n_deg, n_min, n_sec, n_sec1), parse_number) %>% 
	mutate_if(is.numeric, list(~ifelse(is.na(.), 0, .))) %>% 
	mutate(n_sec = ifelse(is.na(n_sec), n_sec1, n_sec)) %>% 
	select(-n_sec1) %>% 
	mutate(lat_dms =  str_glue('{n_deg} {n_min} {n_sec}'),
				 long_dms = str_glue('{e_deg} {e_min} {e_sec}')) %>% 
	# convert from decimal minutes to decimal degrees
	mutate(lat_dd = measurements::conv_unit(lat_dms, 
																			 from = 'deg_min_sec', 
																			 to = 'dec_deg')) %>% 
	mutate(long_dd = measurements::conv_unit(long_dms, 
																				from = 'deg_min_sec', 
																				to = 'dec_deg')) %>% 
	mutate_at(vars(lat_dd, long_dd), as.numeric)

# quick map

library(ggmap)

# bounding box
# 39.534214, 124.159154 ... 39.657415, 129.138604
# 33.694662, 123.491142 ... 34.338454, 130.569658

map <- 
	get_stamenmap(bbox = c(left = 123, 
												 bottom = 33, 
												 right = 	131, 
												 top = 39),
								zoom = 10)

library(ggrepel)
ggmap(map)  + geom_point(data = coords,
						 aes(long_dd ,
						 		lat_dd), 
						 colour = "red",
						 size = 2) +
	geom_label_repel(data = coords,
									aes(long_dd ,
											lat_dd,
											label = site_name),
									size = 2)

# pick points off this map with korea_boundary <- locator()

korea_bb <-  
	data.frame(x = c(125, 
						       130, 
						       130, 
						       125), 
			 y = c(38, 
			 			 38, 
			 			34, 
			 			34))
```

@@

@@

```{r}
#site elevation
mydata_ages <- 
  mydata %>% 
  separate(C14.BP., into = c('age', 'error'),
          sep = "±") %>% 
  mutate(age_ka = parse_number(age) / 1000,
         error = parse_number(error)) %>% 
  mutate(has_sp = ifelse(!is.na(SP.), "yes", "no"))

ggplot(mydata_ages, 
       aes(x = age_ka,
           y = altitude.m._of_main_layer,
       color = as.factor(has_sp))) +
  geom_point(size = 3) +
  xlab("Age of occupation (ka)") +
  ylab("Elevation above sea level (m)") +
  geom_smooth(se = FALSE) +
  scale_colour_discrete(name = "Contains\nstemmed\npoints?") +
  theme_minimal(base_size = 16) 
```

@@

```{r}
#Volume and artefact counts to get density over time. 
kasv_tidy <- 
kasv %>% 
  t %>% 
  as_tibble() %>% 
  setNames(as.character(.[1,])) %>% 
  .[-1,] %>% 
  mutate_all(parse_number) %>% 
  mutate(artefact_density = total_artifacts / volume,
         sites = names(kasv)[-1])

ggplot(kasv_tidy,
       aes(date_age,
           artefact_density)) +
  geom_point(aes(size = total_artifacts)) +
  geom_text_repel(aes(label = sites),
                  nudge_x = 0.25, 
                  nudge_y = 0.25) +
  ylab("Artifact density") +
  xlab("Age of assemblage (ka)") +
  scale_size_continuous(name = "Total number\nof artifacts")  +
  theme_minimal(base_size = 16)

```

@@

```{r}
dates_clean <- 
  dates %>% 
  mutate(age = parse_number(as.character(age)),
         error = parse_number(as.character(error))) %>% 
  filter(age < 50000, method=='AMS')

# devtools::install_github('ahb108/rcarbon')
library(rcarbon)
dates_calibrated <- 
  calibrate(x=dates_clean$age,
               errors=dates_clean$error,
               calCurves='intcal13')

dates_calibrated_spd <-  spd(dates_calibrated, timeRange = c( 45000, 1000))
plot(dates_calibrated_spd)

# Testing Observed SPDs against theoretical models
## recalibrate dates without normalisation to avoid artificial peaks
spd_dates<- calibrate(x=dates_clean$age,errors=dates_clean$error,normalised=FALSE,verbose=F)

spd_test <- modelTest(spd_dates,errors=dates_clean$error,timeRange=c(45000,1000),model='exponential',nsim=200,ncores=3)

summary(spd_test)
plot(spd_test)
```

@@

```{r}
#raw material information
kasr_long1 <- 
  kasr %>% 
  gather(site, 
         count, 
         -X) %>% 
  group_by(site) %>% 
  mutate(percentage = count / sum(count, na.rm = TRUE) * 100,
         total = sum(count, na.rm = TRUE)) %>% 
  filter(!is.na(percentage)) %>% 
  mutate(raw_material = ifelse(percentage >= 10, as.character(X), "other")) %>% 
  mutate(raw_material = ifelse(raw_material == "etc.", "other", raw_material)) %>% 
  select(-X)

  # if percentage is <10%, call it 'other'
kasr_long2 <- 
  kasr_long1 %>% 
  group_by(site, raw_material) %>% 
  summarise(percentage = sum(percentage)) %>% 
  ungroup() %>% 
  left_join(kasr_long1 %>% 
              select(site, raw_material, total),
            by = c('site', "raw_material")) %>% 
  distinct(.keep_all = TRUE) %>%

  # join to get ages of the sites
  left_join(mydata_ages, 
            by = c('site' = 'site_name')) %>% 
  mutate(age_ka = jitter(age_ka),
         age_bin  = ntile(age_ka, 5)) %>% 
  arrange(age_ka) %>% 
  filter(!is.na(age_ka)) %>% 
    mutate(axis_label = glue('{site} ({round(age_ka,1)} ka, n = {total})'))

ggplot(kasr_long2, 
       aes(
           reorder(axis_label, -age_ka),
           percentage,
           fill = raw_material)) +
  geom_col(position = "fill") +
  xlab("Assemblage (youngest at the top)") +
  ylab("Percentage") +
  scale_fill_viridis_d(name = "Raw material type",
                       option = "C") +
  coord_flip() +
  theme_minimal(base_size = 12)

```
Umap

@@

2. Population change
Interpreted in HBE, we expected there might be close connection between environmental change and the appearance of new technology based on the prediction that the new invention was the result of risk. The research suggests that the environmental pressure could be related to population change. We cannot simply predicate that the increased population around 35ka was due to dispersal of Homo Sapiens since there is no reliable human remains in South Korea. However, the migration of Homo sapiens was not impossible scenario at the same time. In fact, one model explains that the emergence of stemmed points and blade industry came from outside of the Korean peninsula and the new technology was the result of the continuous influx of modern human populations from Siberia while the earlier coarse flake trandition was come from southern China [@Bae_2010]. Since this migration model barely explains about preexisted stone artifact assemblages and people in the Korean Peninsula, it is hard to accept his argument without knowing about the native people. In short, we assume that the stemmed points appeared related to the increased population and it can be related to the modern human dispersal or indigenously increased occupants.

@@

------------NOT WORKING-------------

```{r}
#the artifact composition of each sites to figure out what people did as they were occupying the sites.
kasa_long <-
  kasa %>%
  gather(site, 
         count, 
         -X) %>% 
  filter(!is.na(count)) %>% 
  group_by(site) %>% 
  mutate(percentage = count / sum(count, na.rm = TRUE) * 100,
         total = sum(count, na.rm = TRUE)) %>% 
  filter(!is.na(percentage)) %>% 
  # if percentage is <10%, call it 'other'
  mutate(artefact_type = ifelse(as.character(X) == 'stemmed_point',
                                'stemmed_point', 
                                ifelse(percentage >= 10, 
                                as.character(X), "other_tools"))) %>% 
  mutate(axis_label = glue('{site}\n(n = {total})'))  %>% 
  # join to get ages of the sites
  left_join(mydata_ages, 
            by = c('site' = 'site_name')) %>% 
  arrange(age_ka) %>% 
   mutate(age_ka = jitter(age_ka))

ggplot(kasa_long, 
       aes(age_ka, 
           percentage,
           fill = artefact_type)) +
  geom_col() +
  ylab("Percentage") +
  xlab("Age of assemblage (ka)") +
  labs(fill="Artifact type")+
  theme_minimal(base_size = 16)

```

Get the lithic data for some sites:

```{r}
# we'll use googlesheets package to get the data more efficiently
library(googlesheets)

# here is the key to the google sheet: this ggole sheet is same as kasa. so I could do kas_sites <- kasa 
kas_sites <- gs_key("14G0Mp-YSsKBOBPF1Z1zPWQbZQ-sjJSWLetDIQSlbsjw")

# get the tab names from this sheet: But that I couldn't use gs_ws_ls
kas_sites_names <- gs_ws_ls(kas_sites)

# sheets we need here
sheets <- c( "general_info",    "assemblage_info" )

# get the data from the tabs in this sheet!
kas_sites_sheets <- 
  map(sheets, ~kas_sites %>% gs_read_listfeed(ws = .x))
```

Compute artefact volumetric density by frequency of retouched piece (log scale), following Julien Riel-Salvatore and C. Michael Barton in  https://www.jstor.org/stable/4128419

```{r}
# put the site names on this list of data frames
site_names <-  kas_sites_sheets[[1]] %>%  
  select(-X1) %>% 
  names()

# density of arefacts at each site
artefact_density <- 
kas_sites_sheets[[1]] %>% 
  t %>% 
  as.data.frame() %>% 
 rownames_to_column() %>% 
  mutate_all(as.character)

names(artefact_density) <- as.character(as.vector(artefact_density[1, ]))
artefact_density <- artefact_density[-1,]

artefact_density <- 
artefact_density %>% 
  mutate(density = as.numeric(total_artifacts) / as.numeric(volume)) %>% 
   mutate_at(2:ncol(.), as.numeric)

# frequency of retouched artefacts
retouched_artefacts <- 
 kas_sites_sheets[[2]] %>% 
   filter(!X1 %in% c("debris", 
                     "unkown", 
                     "pebble",  
                     "metate", 
                     "flake",
                     "unfinished",
                     "ground",
                     "blank"
                     )) %>% 
    t %>% 
  as.data.frame() %>% 
 rownames_to_column() %>% 
  mutate_all(as.character)

names(retouched_artefacts) <- as.character(as.vector(retouched_artefacts[1, ]))
retouched_artefacts <- retouched_artefacts[-1,]

retouched_artefacts <- 
retouched_artefacts %>% 
  mutate_at(2:ncol(.), as.numeric)  %>% 
  mutate(total_retouched = rowSums(.[,-1], na.rm = TRUE))

# join together the site age, artefact density df with the retouched artefacts df
kas_density_retouch_ages <- 
artefact_density %>% 
  left_join(retouched_artefacts) %>% 
  mutate(proportion_retouched = total_retouched / total_artifacts) %>% 
  mutate(sites = X1)

# compute correlation
kas_sites_retouch_density_corr <- 
cor.test(kas_density_retouch_ages$proportion_retouched,
         kas_density_retouch_ages$density)

r_value <- unname(kas_sites_retouch_density_corr$estimate)

```

```{r}
size <-  8

ggplot(kas_density_retouch_ages,
       aes(density,
           proportion_retouched
           )) +
  geom_point(aes(colour = date_age / 1000,
                 size = total_artifacts,
                 shape = is.na(stemmed_point))) +
 # geom_text_repel(aes(label = sites),
 #                 nudge_x = 0.075, 
 #                 nudge_y = -0.075,
 #                 size = 5) +
  geom_smooth(alpha = 0.2) +
  scale_y_log10(limits = c(0.001, 1),
                labels = scales::comma_format(accuracy = 0.001)) +
  scale_x_log10(limits = c(0.001, 10),
                labels = scales::comma_format(accuracy = 0.001)) +
  theme_minimal(base_size = 12) +
  scale_color_viridis_c(name = "Age of\nassemblage (ka)") +
  xlab("Artifact volumetric density") +
  ylab("Proportion of retouched pieces") +
  annotate("text",
           x = 0.005,
           y = 0.7,
           label = "Curated",
           size = size -2,
           colour = "grey50") +
    annotate("text",
           x = 5,
           y = 0.01,
           label = "Expedient",
           size = size -2,
           colour = "grey50") +
    annotate("text",
           x = 0.01,
           y = 0.01,
           label = glue('r = {round(r_value, 3)}'),
           size = size - 2,
           colour = "grey50")

```

```{r}
# I wanna make simple table for artifact density and frequency of retouched pieces again. Because the complex table above is hard to make comments..

ggplot(kas_density_retouch_ages,
       aes(density,
           proportion_retouched
           )) +
  geom_point(aes(colour = date_age / 1000,
                 size = 7)) +
 # geom_text_repel(aes(label = sites),
 #                 nudge_x = 0.075, 
 #                 nudge_y = -0.075,
 #                 size = 5) +
  geom_smooth(alpha = 0.2) +
  scale_y_log10(limits = c(0.001, 1),
                labels = scales::comma_format(accuracy = 0.001)) +
  scale_x_log10(limits = c(0.001, 10),
                labels = scales::comma_format(accuracy = 0.001)) +
  theme_minimal(base_size = 12) +
  scale_color_viridis_c(name = "Age of\nassemblage (ka)") +
  xlab("Artifact volumetric density") +
  ylab("Proportion of retouched pieces") +
  annotate("text",
           x = 0.005,
           y = 0.7,
           label = "Curated",
           size = size -2,
           colour = "grey50") +
    annotate("text",
           x = 5,
           y = 0.01,
           label = "Expedient",
           size = size -2,
           colour = "grey50") +
    annotate("text",
           x = 0.01,
           y = 0.01,
           label = glue('r = {round(r_value, 3)}'),
           size = size - 2,
           colour = "grey50")

```

Artefact volumetric density and frequency of retouched pieces can be proxies to identify the characteristics of the hunter-gatherers who carried that stone tools whether they are residentially mobile or logistically orgnized foragers based on the premise that the relationship between lithic assemblage composition and land-use strategy represent an expression of economizing behavior.

The artefact volumetric density is defined as the total number of artefacts per cubic meter of excavated sediment, and served as a proxy for accumulation ratio of artefacts. the relative frequency of retouched pieces is the count of retouched flake tools devided by the total number of artifacts.

The two ends of the curated-expedient show that residential to logistical mobility. Expedient assemblages can be depicted as "base camps" or "residences", while curated ones represent either residential mobility, where the central locus of group changes frequently, or certain task groups from a central residential site.

The plot show that younger sites are likely to be used as base camp than older site, which means that people tended to stay longer at one place after around 40-35ka.

(Riel-Salvatore and Barton 2004)

```{r}
# dating
library(rcarbon)

# I'm working through https://cran.r-project.org/web/packages/rcarbon/vignettes/rcarbon.html

kas_ages <- 
kas_density_retouch_ages %>% 
  select(sites, date_age, date_error) %>% 
  mutate(sites = str_remove_all(sites, "_"))

kas_caldates <- 
  calibrate(x = kas_ages$date_age,
            errors = kas_ages$date_error,
            calCurves='intcal13',
            ncores=3) #running calibration over 3 cores

kas_spd = spd(kas_caldates, timeRange = c(50000,25000)) 
plot(kas_spd)

binsense(x=kas_caldates,
         y=kas_ages$sites,
         h=seq(0,5000,100),
         timeRange=c(50000,25000))

ggplot(kas_spd$grid,
       aes(calBP,
           PrDens)) +
  geom_line() +
  scale_y_log10(name = "Summed probability")

# test how the SPD fits to an exponential growth model:
expnull <- modelTest(kas_caldates, 
                     errors=kas_ages$date_error, 
                     nsim=100, 
                     timeRange=c(50000,25000),
                     model="uniform",
                     runm=10)

plot(expnull)

```
